Using VariantSpark to learn population structure from genomic profiles

### 2015-10-26 doc/publications/Recomb15/recomb15.tex
broken thing

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}
\usepackage{placeins}
\usepackage{hyperref}
%

@@

\maketitle
%
\begin{abstract}
This paragraph shall summarize the contents of the paper
in short terms.
\end{abstract}
%
\section{Introduction}
%

@@

\marginpar{I don't understand this. Can't you just sak a patient about their ancestry?}
Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
information~\cite{Zheng2014}. Such clinical application are severely hampered for patients where accurate population
association is unknown due to a diverse migrational background. Determining the population association from genomic
information may provide vital insights for treatment and diagnostic in such cases.

@@

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
association and compare our results to the ones obtained using deep learning.
%\marginpar{TODO Rob can you think of a better description of what it does}
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
by fitting a linear combination of the feature variables.  It is therefore able to adjust the weight of the individual
genotypes to increase the accuracy of the multi-class prediction.

@@

To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF
method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first
section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
DL method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
\marginpar{TODO elaborate on the error Aidan: See near end of paragraph} 
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
along with the standard error of the mean (SEM). We also use the Adjusted Rand Index (ARI) as a metric for model quality
as it allows us to compare the classification results with \kMeans{} clustering models. We also calculate the error
for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
resulting models along with the standard error.

\marginpar{TODO elaborate on accuracy measures Aidan: See above paragraph}
As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately, with an average ARI over ten models of \(0.596 \pm 0.067\), see table~\ref{RF}.
%Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by the LR, with an ARI of \(0.923 \pm 0.015\).
This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.

\marginpar{not sure what is meant here. RF, LR etc are not multivariate -- there is only one $y$ variable, obesity. They
are multi-variable.}

@@

These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.

\FloatBarrier
\clearpage
\marginpar{TODO report bd's accuracy measure as the main measure (add ARI as the additional information linking to our
  paper) Aidan:See email}
\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether another multivariate approach, DL, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
Scotland).

The mean ARI for our \kMeans clustering models on this dataset is \(0.960 \pm 0.004 \). 
%DENIS: lower compared to what ?
Although slightly lower, this value is similar to the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual is misclassified as CHB). 
Our supervised logistic regression approach, however, improves upon this result and achieves an ARI of 1 across the ten folds.

%DENIS: I don't understand this 
To reiterate, these results are from the three populations, ASW, CHB and GBR, that belong to the super-populations, AFR,
EAS and EUR, respectively.  As we mention in the introduction, we observed a perfect clustering result when we previously clustered
 these three super-populations (rather than one population from each), with an ARI of 1.0. However, when
we introduce the fourth super-population into the mix, AMR, the ARI drops to 0.84.

Given the perfect result when building
the logistic regression model on this small subset of data, and the similarity with the other models,
we expand our dataset to include every population from the 1000 Genomes Project phase1 dataset.

@@

To determine where the model is underperforming, we also record the precision and recall for each of the groups.
We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
This demonstrates that each logistic regression model was able to classify every EAS individual correctly, while not misclassifying any other individuals as EAS.
For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
Consequently, this leds to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
Although these misclassifications as AMR are relatively frequent, we did not observe any AFR individuals being classified as EUR, or {\it vice versa}, in any of the ten folds.
Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
These results demonstrate the diversity, or relative difference of some of the groups.

\marginpar{TODO [Discuss supervised vs unsupervised] Aidan: wrote in previous section (where I mention initialisation)} \marginpar{TODO can you drill down to the population clusters and
  see if and where the best performing method still makes mistakes - this will help in the conclusion to argue that
  actually the labels are wrong rather than the method not being able to capture the structure (e.g. say if AMR are
  mixed in with GBR TSI) Aidan: not in time :( }

@@

Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests.
This is likely due to RF being an ensemble method. Where logistic regression and \kMeans{} clustering build a single model,
RF is building a forest of trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.

@@

\bibliographystyle{my_plain}
\marginpar{FIXED:  TODO you need to find a way of shortening the Locket paper author list}
\bibliography{recomb}

>>>>

% This is LLNCS.DOC the documentation file of
% the LaTeX2e class from Springer-Verlag
% for Lecture Notes in Computer Science, version 2.4
\documentclass{llncs}
\usepackage{llncsdoc}
\usepackage{placeins}
\usepackage{hyperref}
\usepackage{amsmath}
%

@@

\maketitle
%
\begin{abstract}
Using our 
We apply various machine learning algorithms to the 1092 individuals in the 1000 Genomes Project dataset.
Although each algorithm we investigate can cluster individuals from three populations with near perfect accuracy,
we find logistic regression to be superior. Comparing  With the entire dataset,

\end{abstract}
%
\section{Introduction}
%

@@

\marginpar{I don't understand this. Can't you just ask a patient about their ancestry?}
Determining a person's ethnicity has become important for medical applications, such as HLA allele genotyping from SNP
information~\cite{Zheng2014}. Such clinical application are severely hampered for patients where accurate population
association is unknown due to a diverse migrational background. Determining the population association from genomic
information may provide vital insights for treatment and diagnostic in such cases.

@@

In this paper we therefore systematically evaluate supervised and unsupervised methods for determining population
association and compare our results to the ones obtained using deep learning.
%\marginpar{TODO Rob can you think of a better description of what it does}
One of the simplest supervised machine learning methods is logistic regression (LR), which creates a decision boundary
by fitting a linear combination of the feature variables.  In this appliction it is not fesible to model interactions,
so the model is retricted to main effects.

@@

To demonstrate \variantSpark's capability, we determine the cross-validation accuracy of the supervised multivariate RF
method and compare it against the simple supervised method LR as well as the unsupervised \kMeans{} in the first
section.  In the second section we compare LR and \kMeans{} against the reported accuracy of the supervised multivariate
deep learning (DL) method.  In section three we record the accuracy of LR and \kMeans{} on the full dataset set using all populations.
In the last section we discuss the memory consumptions and runtime of LR compared to \kMeans{}.

\clearpage
\hrule
I think the use of the term ``multivariate'' is a bit loose. We have only one $y$ variable -- obesity. By analogy with linear regression:
\begin{itemize}
\item simple linear regression $y=\beta_0 + \beta_1x_1$
\item multi-variable regression $y=\beta_0 + \beta_1x_1 + \beta_2 x_2$
\item multivariate regression $\underset{n\times m}{Y}= \underset{(n\times p+1)}{X} \underset{(p+1 \times m)}{B}$
\end{itemize}
So RF, DL and LR are all multi-variable (in this application).

A multivarite technique we could use is PCA, $T = X W$ where $W$ is a $p \times p$ matrix whose columns are the eigenvectors of $X^tX$. 
\hrule

@@

\subsubsection{Random Forest does not scale to whole genome analysis and is outperformed by logistic regression}
In this section we compare RF against \kMeans\ and the LR.  
Despite using {\sc Spark} and {\sc MLlib}, RF is not able to scale to the full dataset of 1000 individuals and 40
Million variables, due to a limitation of the Java virtual machine (maximum array size) that causes {\sc Mllib}'s
random forest implementation to fail. This error is due to the number of features present in each sample,
with a potential solution being to limit the features through feature selection, or simply using a subset of the dataset.
We therefore use only chromosome 22 (494,328 alleles) for the comparison in this section.  We train a RF
and LR on this subset of genomic profiles from the 1000 Genomes data using the annotated labels of all
super-populations.  We perform 10-fold cross-validation and report the mean error over all training and testing folds
along with the standard error of the mean (SEM). We also use the Adjusted Rand Index (ARI) as a metric for model quality
as it allows us to compare the classification results with \kMeans{} clustering models. We also calculate the error
for the two classification methods, which is simply the number of misclassified individuals divided by the number of samples.
We perform 10 runs of \kMeans{} with 4 clusters (\(k=4\)) and report the mean ARI of the
resulting models along with the standard error.

As reported in our earlier work, \kMeans\ is not able to fit the dataset accurately, with an average ARI over ten models of \(0.596 \pm 0.067\), see table~\ref{RF}.
%Although capable of fitting a relatively accurate model (\(ARI=0.837\)), The average ARI over ten models is  \(0.596 \pm 0.067\), see table~\ref{RF}.
RF achieves a better fitting model with a mean ARI of \(0.769 \pm 0.021\), however it is outperformed by the LR, with an ARI of \(0.923 \pm 0.015\).
This indicates that while applying a supervised learning method results in higher accuracy, a multivariate approach may not be needed to accurately fit ethnicity.

@@

%These results also demonstrate how much the quality of the models can vary between runs (or folds) for supervised and unsupervised learning.
%For both of the supervised methods, RF and LR, the ARI was fairly consistent when testing on each holdout set, with a standard deviation of
%\( 0.064 \) and 0.048, respectively. For the \kMeans{} models however, the standard deviation was 0.211.
%One pitfall of \kMeans{} clustering is that depending on the starting points, the resulting model may represent a local optimum of the dataset~\cite{Steinley2003}.
%{\sc MLlib} uses {\sc \kMeans\textbar\textbar}~\cite{Bahmani2012} (a parallel implementation of \kMeans{}++) for the cluster initialisation.
%Although this produces close to the optimal solution for the initial cluster centres, due to the heuristical nature of the algorithm,
%these initial cluster centres will be different for each model. Due to the amount of variance in the 1000 Genomes dataset, it is possible that multiple local
%optima are close to the global optimum, and the algorithm is choosing one that does not represent population structure.

\subsubsection{Logistic regression and \kMeans{} capture the population better than deep learning in a three population test}
In this section we investigate whether another multivariate approach, DL, is able to achieve a better result than
RF by again comparing it against \kMeans{} and logistic regression.  We compare against the reported accuracy of the
deep learning approach using the {\sc ADAM} framework~\cite{Massie2013} as reported on their webpage~\cite{Ferguson}.
As the reported results were performed on only three of the 14 populations, we hence restrict our comparison also to ASW
(Americans of African Ancestry in SW USA), CHB (Han Chinese in Beijing, China), and GBR (British in England and
Scotland).

The mean ARI for our \kMeans{} clustering models on this dataset is \(0.960 \pm 0.004 \). 
This value is slightly less than the one reported for the deep-learning approach, which achieved an ARI of 0.976 (where one ASW and one GBR individual are misclassified as CHB). 
Our supervised logistic regression approach, however, improves upon this result and achieves an ARI of 1 across the ten folds.

Although we see a slight improvement with logistic regression in regards to other algorithms, their results are still very similar.
However, there is not much room in the first place for logistic regression to demonstrate any potential increases in model accuracy.
We therefore proceed to apply our method to the entire 1000 Genomes Project phase 1 dataset, which we previously clustered with an ARI of 0.84.

@@

To determine where the model is underperforming, we also record the precision and recall for each of the groups.
We observe that for each model, the precision and recall are always 1.0 for the EAS super-population.
This demonstrates that each logistic regression model was able to classify every EAS individual correctly, while not misclassifying any other individuals as EAS.
For the AMR group, however, we consistently observed a relatively low precision (\(0.829 \pm{} 0.02 \)).
This is due to individuals from the AFR and EUR super-populations being misclassified as AMR.
Consequently, this leads to these two super-populations having a lower recall of \(0.956 \pm 0.01 \) and \(0.950 \pm 0.02 \), respectively.
Although these misclassifications as AMR are relatively frequent, we did not observe any AFR individuals being classified as EUR, or {\it vice versa}, in any of the ten folds.
Also, only in one fold did we observe one AMR individual misclassified as another super-population (AFR).
These results demonstrate the diversity, or relative difference of some of the groups.

@@

Time and resources vary greatly between the three algorithms, with the most resource intensive being random forests.
This is likely due to RF being an ensemble method. Where logistic regression and \kMeans{} clustering build a single model,
RF is building an forest of decision trees. So not only is training longer, but then each sample in the test fold must be classified by each and every tree.

@@

\bibliographystyle{my_plain}
\bibliography{recomb}
